Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-09-25 paper/appendix.tex

\newpage
\appendix
\section{Appendix}
\subsection{Feature importance}

>>>>

\newpage
\appendix
\section{Appendix}

\subsection{Proof of consistency theorem}
\label{appendix:theorem_proof}

Proof of the theorem in \ref{sec:theorem}:

\begin{theorem*}[B2B consistency - general case]

     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full rank noise.
     %
     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimize
     %
     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $E\hat H = \hat H$
\end{theorem*}

\begin{proof}

 Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B, we have
 \begin{align*}
    \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
%                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2]\\
                                                        &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
     \label{eq:doublenorm}
\end{align*}
\begin{align*}
    \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
%    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]\\
    &= EF \hat G
\end{align*}

Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero. Let $Z=F^\dagger EF\hat G$. We have $FZ = FF^\dagger EF  \hat G= EF\hat G$ ($FF^\dagger E =E$ since $F$ is full rank on $Img(E)$). Since E is a contraction, we have $ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$. Therefore,
 $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$

But as $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, the above inequality is an equality, $Z=\hat G$ and $EF\hat G = F\Hat G$. Therefore, $\left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 = \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 = \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2$ is minimised by $\hat G$ and $\hat H$. Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$. This completes the proof.
\end{proof}

\newpage
\subsection{Feature importance}
\label{appendix:feature_importance}

### 2019-09-25 paper/meg_exps.tex

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_evoked.pdf}
  \caption{102 subjects read $\approx$ 2,700 words while their brain activity was recorded with MEG. Top. Average brain response to words relative to their onset (gradiometers), as viewed from above the head. Bottom. Each line represents a MEG sensor (magnetometer), color-coded by spatial position. Posterior responses, typical of primary visual cortex activity, peak around 100 ms after word onset and are followed by an anterior propagation of activity typical of semantic processing in the associative cortices. Vertical lines indicate moments of above topographies.
  }
  \label{fig:meg_evoked}
\end{figure}

@@

When human subjects read words, the number of letters is known to modulate
brain activity from 100ms after word onset \citep{pegado2014timing}.
Similarly, word frequency is known to modulate brain activity around 400ms
\citep{kutas2011thirty}. In the present study, we consider these two well-known features, as well as an untested feature ('Word Function') and dummy variable.
%
"Word Length' refers to the total number of letters, and is expected to specifically cause a variation in the early evoked MEG responses (i.e. from 100 ms after stimulus onset) elicited by the retinotopically-tuned visual cortices.
%
'Word Frequency', as derived with WordFreq \citep{speerwordfreq}, indexes how frequently each word appears in Dutch (using the Zipf logarithmic scale of \citep{van2014subtlex}). This feature is expected to specifically cause a variation in the late evoked MEG responses (i.e. from 400 ms), because it variably engages semantic processes in the temporal cortices \citep{kutas2011thirty}.
%
Finally, 'Word Function' indicates whether each word is a content word (i.e. a noun, a verb, an adjective or an adverb) or a function word (i.e. a preposition, a conjunction, a determinant, a pronoun or a numeral), and was derived from Spacy's part of speech tagger \citep{spacy2}. To our knowledge, this feature has not been thouroughly investigated with MEG. Its causal contribution to reading processes in the brain is thus largely unknown.
%
To verify that B2B and other models would not inadequately identify non-causal features, we added a dummy feature, constructed from a noisy combination of "Word Length" and "Word Frequency":
\begin{equation}
  dummy = z(length) + z(frequency) + \mathcal{N}
\end{equation}
where $z$ transforms features into zero-mean and unit variance vectors and $\mathcal{N}$ is random vectors sampling a zero-mean unit-variance Gaussian distribution.

@@

We compare B2B to four standard models: forward regression, backward regression, CCA and PLS, as implemented in scikit-learn, and optimized with nested cross validation over twenty l2 regularization parameters logarithmically spaced between $10^-4$ and $10^4$ or 1 to 4 caconical components.

We used the feature importance described in \ref{algorithm:b2b_fi} to assess the extent to which each feature $X_i$ specifically improves the prediction of held-out $Y$ data.

Each model was implemented for each subject and each time sample independently. Pairwise comparison across models were performed using a wilcoxon test on the $\Delta R$ averaged across time.

Corresponding results can be found in Figure \ref{fig:meg_results}. Significant pairwise differences across models are highlighted with the top horizontal bars in \ref{fig:meg_results}.

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg.pdf}
  \caption{Forward regression (Blue), PLS (orange), CCA (green) and B2B (red) are compared on their ability to reliably predict single-trial MEG signals evoked by words. Left. Average improvement of correlation coefficient $\Delta R$ obtained on held-out data when each of the four features "Word Length", "Word Frequency", "Word Function" and "Dummy Variable" was provided to the model \ref{algorithm:b2b_fi}. Error bars indicate standard error of the mean (SEM) across subjects. Right. Average $\Delta R$ across time. Each dot represents a subject. The top horizontal lines indicate significant differences between pairwise comparison across models, color coded by the winning model. B2B outperforms all models on meaningful features, at the exception of CCA for the "word length".}
  \label{fig:meg_results}
\end{figure}

@@

First, \ref{fig:meg_supp} shows that the Backward model decodes the dummy variable well above chance level. In addition, "Word Length" and  "Word Frequency" present a similar decoding time course, even though these features are known to influence early and late MEG responses respectively \citep{kutas2011thirty}. These results illustrate the well-known fact that backward modelling cannot be used to estimate the causal contribution of collinear features.

We thus focus on valid modeling (Forward Regression, PLS, CCA, and B2B) and estimate the change of correlation coefficients $\Delta R$ induced by the introduction of each feature into the model \ref{algorithm:b2b_fi}. As expected, none of the models significantly predicted the 'Dummy Variable' to be a causal factors (i.e. $\Delta R < 0$ for all models, all $p > .089$)

@@

The results show that all models significantly detect that both word length and
word frequency modulate brain responses after word onset. However,
forward regression detected word length ($t$-value$=10.6$, $p=10^{-16}$) and
word frequency ($t$-value$=11.1$, $p=10^-17$) less reliably than B2B (length:
$t$-value$=18.7$, $p=10^{-31}$, frequency: $t$-value$=15.9$, $p=10^{-26}$). B2B
was less sensitive than the backward regression (length: $t$-value$=33.6$,
$p=10^{-50}$, frequency: $t$-value$=31.7$, $p=10^{-48}$). However, as backward models confound collinear factors, this high sensitivity likely comes from a loss of specificity. This possibility is corroborated by the fact that the effects of word length and word frequency identified with backward regression present largely similar dynamics.

Finally we investigate in Figure \ref{fig:megresult} whether B2B was sufficiently sensitive to reliably detect a multiplicity of subtle features, such as individual letters and part-of-speech. The vast majority of the visual and lexical features appeared to be reliably detected (see Figure 4 in Supplementary Material). B2B thus promises to improve the detectability of subtle effects.

>>>>

\begin{figure}[t!]
  \begin{minipage}[c]{0.6\textwidth}
    \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_evoked.pdf}
  \end{minipage}\hfill
  \begin{minipage}[c]{0.4\textwidth}
    \caption{Ninety subjects read approximately 2,700 words while their brain activity was recorded with MEG. Top. Average brain response to words (word onset at t=0 ms), as viewed from above the head (red= higher gradient of magnetic flux). Bottom. Each line represents magnetometer, color-coded by spatial position. Posterior responses, typical of primary visual cortex activity, peak around 100 ms after word onset and are followed by an anterior propagation of activity typical of semantic processing in the associative cortices.
    }
    \label{fig:meg_evoked}
  \end{minipage}
\end{figure}

@@

We are here interested in identifying the features of word that modulate (i.e. cause) brain activity. We consider four distinct, but colinear features.
%
"Word Length' refers to the total number of letters. Word Length is expected to specifically cause a variation in the early evoked MEG responses (i.e. from 100 ms after stimulus onset) elicited by the retinotopically-tuned visual cortices (e.g. \citep{pegado2014timing}.).
%
'Word Frequency', as derived with WordFreq \citep{speerwordfreq}, indexes how frequently each word appears in Dutch (using the Zipf logarithmic scale of \citep{van2014subtlex}). Word Frequency is expected to specifically cause a variation in the late evoked MEG responses (i.e. from 400 ms), because it variably engages semantic processes in the temporal cortices \citep{kutas2011thirty}.
%
'Word Function' indicates whether each word is a content word (i.e. a noun, a verb, an adjective or an adverb) or a function word (i.e. a preposition, a conjunction, a determinant, a pronoun or a numeral), and was derived from Spacy's part of speech tagger \citep{spacy2}. To our knowledge, this feature has not been thouroughly investigated with MEG. Its causal contribution to reading processes in the brain thus remains unclear.
%
Finally, to verify that B2B and other methods would not inadequately identify non-causal features, we added a dummy feature, constructed from a noisy combination of "Word Length" and "Word Frequency":
\begin{equation}
  dummy = z(length) + z(frequency) + \mathcal{N}
\end{equation}
where $z$ transforms features into zero-mean and unit variance vectors and $\mathcal{N}$ is a random vector sampling a zero-mean unit-variance Gaussian distribution.

@@

We compare B2B to four standard methods: Forward regression, Backward regression, CCA and PLS, as implemented in scikit-learn \citep{scikit}, and optimized with nested cross validation over twenty l2 regularization parameters logarithmically spaced between $10^-4$ and $10^4$ (for regression methods) or 1 to 4 canonical components (for cross-decomposition methods).

We used the feature importance described in Algorithm \ref{algorithm:b2b_fi} to assess the extent to which each feature $X_i$ specifically improves the prediction of held-out $Y$ data.

Each model was implemented for each subject and each time sample independently. Pairwise comparison between models were performed using a Wilcoxon test across subjects (n=90) using the average $\Delta R$ across time.

The chapter corresponding results can be found in Figure \ref{fig:meg_results}. Significant ($p<0.05$) pairwise differences across models are highlighted with the top horizontal bars in \ref{fig:meg_results}.

\begin{wrapfigure}{r}{.5\textwidth}
  \begin{center}
    \includegraphics[width=0.48\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg.pdf}

    \label{fig:meg_results}
  \end{center}
  \caption{Multiple models (color-coded) are compared on their ability to reliably predict single-trial MEG signals evoked by words. Left. Average improvement of correlation coefficient $\Delta R$ for each pf the four features (rows). Error bars indicate standard error of the mean (SEM) across subjects. Right. Average $\Delta R$ across time for each subject (dot). Top horizontal lines indicate when B2B significantly outperform other methods.}
\end{wrapfigure}

@@

Supplementary Figure \ref{fig:meg_supp} shows that the Backward model decodes the dummy variable well above chance level. In addition, "Word Length" and "Word Frequency" present a similar decoding time course, even though these features are known to specifically influence early and late MEG responses respectively \citep{kutas2011thirty}. These results illustrate the well-known fact that backward modelling cannot be used to estimate the causal contribution of collinear features.

We thus focus on the valid methods (i.e. Forward Regression, PLS, CCA, and B2B) and estimate the change of correlation coefficients $\Delta R$ induced by the introduction of each feature into the model \ref{algorithm:b2b_fi}. As expected, none of the models significantly predicted the 'Dummy Variable' to be a causal factors (i.e. $\Delta R < 0$ for all models, all $p > .089$)

@@

On average across time, "Word Length" and "Word Frequency" improved the prediction performance of all methods $\Delta R>0$ (all $p<0.0001$). As expected, the time course of "Word Length" and "Word Frequency" rose from $\approx$ 100 ms and from $\approx$ 400 ms respectively. Furthermore, 'Word Function' improved the prediction performance of all models (all $p < 0.0002$) except PLS ($p=0.7989$). Overall, these results confirm that "Word Length", "Word Frequency" and "Word Function" each causally influence specific periods of brain responses to words.

To assess which model would be more sensitive to these causal discoveries, we performed pairwise comparisons between models and across subjects (Figure \ref{fig:meg} right). For "Word Length" B2B outperforms all models (all $p < 0.00001$) but CCA ($p=0.0678$). For "Word Frequency", B2B outperforms all models (all $p < 0.0006$). For "Word Function", B2B outperforms all models (all $p < 0.0015$). Overall, these results suggest that B2B reliably outperform standard methods, especially when the effects are difficult to detect.
